---
title: 'Group X Final Project '
output:
  html_notebook: default
  pdf_document: default
Dataset: "https://www.kaggle.com/datasets/daveianhickey/2000-16-traffic-flow-england-scotland-wales"
---

# **Analyzing Traffic Accidents: Patterns, Trends, and Predictive Insights**

## Importing Datasets and Libraries

```{r}
library(randomForest) 
library(Rborist) 
library(rpart)  
library(rpart.plot) 
library(dplyr) 
library(vip)  
library(foreach) 
library(doParallel) 
library(caret)
library(dplyr)
library(lubridate)
library(ggplot2)
```

```{r}
# Read each dataset 
data1 <- read.csv("accidents_2005_to_2007.csv", stringsAsFactors = FALSE) 
data2 <- read.csv("accidents_2009_to_2011.csv", stringsAsFactors = FALSE) 
data3 <- read.csv("accidents_2012_to_2014.csv", stringsAsFactors = FALSE)   
# Combine them into one dataframe 
data <- rbind(data1, data2, data3)   
rm(data1) 
rm(data2) 
rm(data3)
```

```         
```

### Data Cleanup

```{r}
dataTime <- data[,c("Date","Day_of_Week","Time")]

cols_to_keep <- c(
 "Police_Force", "Accident_Severity", "Number_of_Vehicles","Number_of_Casualties", "Road_Type", "Speed_limit", "Junction_Control",
  "Pedestrian_Crossing.Human_Control", "Pedestrian_Crossing.Physical_Facilities",
  "Light_Conditions", "Weather_Conditions", "Road_Surface_Conditions",
  "Special_Conditions_at_Site", "Carriageway_Hazards", "Urban_or_Rural_Area",
  "Did_Police_Officer_Attend_Scene_of_Accident"
)

data <- data[, cols_to_keep]


categorical_cols <- c(
"Accident_Severity", "Road_Type", "Speed_limit", "Junction_Control","Pedestrian_Crossing.Human_Control", "Pedestrian_Crossing.Physical_Facilities","Light_Conditions", "Weather_Conditions", "Road_Surface_Conditions","Special_Conditions_at_Site", "Carriageway_Hazards", "Urban_or_Rural_Area","Did_Police_Officer_Attend_Scene_of_Accident"
)

data[categorical_cols] <- lapply(data[categorical_cols], as.factor)
```

```{r}
summary(data)
```

```{r}
missing_counts <- colSums(is.na(data)) 
missing_counts
```

### Research Question 1: Predicting Accident Severity with road conditions

#### Random Forest Regression

```{r}
run_rf_analysis <- function(data, mtry, ntree, exclude_cols = NULL) {
  #' Run Random Forest analysis with fixed sampling parameters
  #'
  #' @param data Input data frame (must contain 'Accident_Severity' column)
  #' @param mtry Number of variables tried at each split
  #' @param ntree Number of trees to grow
  #' @param exclude_cols Vector of column names to exclude from predictors
  
  # Set fixed parameters
  sample_fraction <- 0.01
  train_frac <- 0.7
  target_var <- "Accident_Severity"
  
  # Sample data
  reduced_data <- data[sample(nrow(data), size = round(sample_fraction * nrow(data))), ]
  
  # Split into train/test
  index <- sample(1:nrow(reduced_data), round(nrow(reduced_data) * train_frac))
  train <- reduced_data[index, ]
  test <- reduced_data[-index, ]
  
  # Remove excluded columns from predictors
  if(!is.null(exclude_cols)) {
    predictors <- setdiff(names(train), c(target_var, exclude_cols))
    formula <- as.formula(paste(target_var, "~", paste(predictors, collapse = "+")))
  } else {
    formula <- as.formula(paste(target_var, "~ ."))
  }
  
  # Build RF model
  rf_model <- randomForest(formula,
                         data = train,
                         mtry = mtry,
                         ntree = ntree,
                         importance = TRUE)
  
  # Model evaluation
  test_predictions <- predict(rf_model, newdata = test)
  conf_matrix <- confusionMatrix(test_predictions, test[[target_var]])
  
  # Return results as list
  return(list(
    model = rf_model,
    train = train,
    test = test,
    predictions = test_predictions,
    confusion_matrix = conf_matrix,
    accuracy = conf_matrix$overall['Accuracy'],
    importance = importance(rf_model)
  ))
}
```

```{r}
results1 <- run_rf_analysis(data = data, mtry = 8, ntree = 600, exclude_cols = c("Police_Force"))
results2 <- run_rf_analysis(data = data, mtry = 6, ntree = 600, exclude_cols = c("Police_Force"))
results3 <- run_rf_analysis(data = data, mtry = 10, ntree = 600, exclude_cols = c("Police_Force"))

# Access results:
results1$accuracy         # Test accuracy
results2$accuracy 
results3$accuracy 

```

```{r}
varImpPlot(results1$model)
results1$importance
```

### Research Question 2: Predicting Number of Casualties with road conditions

```{r}
run_casualty_rf <- function(data, mtry, ntree, exclude_cols = NULL, seed = NULL) {
  #' Run Random Forest regression for Number_of_Casualties prediction
  #'
  #' @param data Input data frame
  #' @param mtry Number of variables tried at each split
  #' @param ntree Number of trees to grow
  #' @param exclude_cols Vector of column names to exclude from predictors
  #' @param seed Optional random seed for reproducibility
  
  # Set seed if provided
  if (!is.null(seed)) set.seed(seed)
  
  # Fixed parameters
  sample_fraction <- 0.001
  train_frac <- 0.7
  target_var <- "Number_of_Casualties"
  
  # Sample data
  reduced_data <- data[sample(nrow(data), size = round(sample_fraction * nrow(data))), ]
  
  # Split into train/test
  index <- sample(1:nrow(reduced_data), round(nrow(reduced_data) * train_frac))
  train <- reduced_data[index, ]
  test <- reduced_data[-index, ]
  
  # Remove excluded columns
  if (!is.null(exclude_cols)) {
    predictors <- setdiff(names(train), c(target_var, exclude_cols))
    formula <- as.formula(paste(target_var, "~", paste(predictors, collapse = "+")))
  } else {
    formula <- as.formula(paste(target_var, "~ ."))
  }
  
  # Build RF model
  rf_model <- randomForest(formula,
                          data = train,
                          mtry = mtry,
                          ntree = ntree,
                          importance = TRUE)
  
  # Generate predictions
  test_predictions <- predict(rf_model, newdata = test)
  
  # Calculate metrics
  rmse <- sqrt(mean((test_predictions - test[[target_var]])^2))
  r_squared <- cor(test_predictions, test[[target_var]])^2
  
  # Visualization
  varImpPlot(rf_model, main = "Variable Importance for Casualty Prediction")
  
  # Return results
  return(list(
    model = rf_model,
    train = train,
    test = test,
    predictions = test_predictions,
    rmse = rmse,
    r_squared = r_squared,
    importance = importance(rf_model),
    formula = formula
  ))
}
```

```{r}
casualty_results1 <- run_casualty_rf(data = data,mtry = 4,ntree = 600,exclude_cols = c("Police_Force", "Accident_Severity", "Number_of_Vehicles","Number_of_Casualties", "Road_Type", "Speed_limit", "Junction_Control",
  "Pedestrian_Crossing.Human_Control", "Pedestrian_Crossing.Physical_Facilities",
  "Light_Conditions", "Weather_Conditions", "Road_Surface_Conditions",
  "Special_Conditions_at_Site", "Carriageway_Hazards", "Urban_or_Rural_Area"))

casualty_results2 <- run_casualty_rf(data = data,mtry = 8,ntree = 600,exclude_cols = c("Police_Force", "Accident_Severity"))
casualty_results3 <- run_casualty_rf(data = data,mtry = 6,ntree = 600,exclude_cols = c("Police_Force", "Accident_Severity"))

casualty_results1$rmse
casualty_results2$rmse
casualty_results3$rmse
```

```{r}
casualty_results1$r_squared
casualty_results2$r_squared
casualty_results3$r_squared
```

### Research Question 3: Forecasting Police Response

```{r}
run_police_force_rf <- function(data, mtry, ntree, exclude_cols = NULL, seed = NULL) {
  #' Run Random Forest regression for numeric Police_Force prediction
  #'
  #' @param data Input data frame
  #' @param mtry Number of variables tried at each split
  #' @param ntree Number of trees to grow
  #' @param exclude_cols Vector of column names to exclude from predictors
  #' @param seed Optional random seed for reproducibility
  
  # Set seed if provided
  if (!is.null(seed)) set.seed(seed)
  
  # Fixed parameters
  sample_fraction <- 0.01
  train_frac <- 0.7
  target_var <- "Police_Force"
  
  # Sample data
  reduced_data <- data[sample(nrow(data), size = round(sample_fraction * nrow(data))), ]
  
  # Split into train/test
  index <- sample(1:nrow(reduced_data), round(nrow(reduced_data) * train_frac))
  train <- reduced_data[index, ]
  test <- reduced_data[-index, ]
  
  # Remove excluded columns
  if (!is.null(exclude_cols)) {
    predictors <- setdiff(names(train), c(target_var, exclude_cols))
    formula <- as.formula(paste(target_var, "~", paste(predictors, collapse = "+")))
  } else {
    formula <- as.formula(paste(target_var, "~ ."))
  }
  
  # Build RF model
  rf_model <- randomForest(formula,
                         data = train,
                         mtry = mtry,
                         ntree = ntree,
                         importance = TRUE)
  
  # Generate predictions
  test_predictions <- predict(rf_model, newdata = test)
  
  # Calculate regression metrics
  rmse <- sqrt(mean((test_predictions - test[[target_var]])^2))
  
  # Proper R² calculation (1 - SSE/SST)
  SSE <- sum((test[[target_var]] - test_predictions)^2)
  SST <- sum((test[[target_var]] - mean(test[[target_var]]))^2)
  r_squared <- 1 - (SSE/SST)
  
  # Visualization
  varImpPlot(rf_model, main = "Variable Importance for Police Force Prediction")
  
  # Actual vs Predicted plot
  plot(test[[target_var]], test_predictions,
       main = "Actual vs Predicted Police Force",
       xlab = "Actual", ylab = "Predicted")
  abline(0, 1, col = "red")
  
  # Return results
  return(list(
    model = rf_model,
    train = train,
    test = test,
    predictions = test_predictions,
    rmse = rmse,
    r_squared = r_squared,
    importance = importance(rf_model),
    formula = formula
  ))
}
```

```{r}

police_results1 <- run_police_force_rf(data = data,mtry = 4,ntree = 600,exclude_cols = c( "Accident_Severity"))
police_results2 <- run_police_force_rf(data = data,mtry = 8,ntree = 600,exclude_cols = c( "Accident_Severity"))
police_results3 <- run_police_force_rf(data = data,mtry = 6,ntree = 600,exclude_cols = c( "Accident_Severity"))

police_results1$rmse
police_results2$rmse
police_results3$rmse
```

```         
```

### Research Question 2: What time of day sees most accidents

```{r}

library(dplyr)
library(lubridate)
library(tidyr)
library(randomForest)
library(caret)  # For confusionMatrix (accuracy metrics)

# 1. Load and clean data (handle missing DateTime)
dataTime_clean <- dataTime %>%
  mutate(DateTime = as.POSIXct(paste(Date, Time), format = "%d/%m/%Y %H:%M")) %>%
  filter(!is.na(DateTime))  # Drop invalid rows

# 2. Create hourly bins (fill gaps with 0 counts)
hourly_bins <- dataTime_clean %>%
  mutate(DateTime_floor = floor_date(DateTime, unit = "hour")) %>%
  complete(
    DateTime_floor = seq(min(DateTime_floor), max(DateTime_floor), by = "1 hour"),
    fill = list(AccidentCount = 0)  # Treat missing hours as 0 accidents
  )

# 3. Feature engineering
accident_counts <- hourly_bins %>%
  mutate(
    Year = year(DateTime_floor),
    Month = factor(
      month(DateTime_floor, label = TRUE, abbr = FALSE),
      levels = month.name,
      ordered = TRUE
    ),
    Day = day(DateTime_floor),
    Hour = hour(DateTime_floor),
    TimeOfDay = factor(
      case_when(
        Hour >= 5 & Hour < 12 ~ "Morning",
        Hour >= 12 & Hour < 17 ~ "Afternoon",
        Hour >= 17 & Hour < 21 ~ "Evening",
        TRUE ~ "Night"
      ),
      levels = c("Morning", "Afternoon", "Evening", "Night"),
      ordered = TRUE
    )
  ) %>%
  group_by(Year, Month, Day, Hour, TimeOfDay) %>%
  summarise(AccidentCount = n(), .groups = "drop") %>%
  mutate(AccidentCount = ifelse(is.na(AccidentCount), 0, AccidentCount))

# 4. Train-test split (80% train, 20% test)
set.seed(123)
train_frac <- 0.8
index <- sample(1:nrow(accident_counts), round(nrow(accident_counts) * train_frac))
train_data <- accident_counts[index, ]
test_data <- accident_counts[-index, ]

# 5. Train Random Forest
rf_model <- randomForest(
    AccidentCount ~ Hour + TimeOfDay + Month,
  data = train_data,
  importance = TRUE,
  ntree = 500
)

# 6. Evaluate training accuracy
train_pred <- predict(rf_model, train_data)
train_rmse <- sqrt(mean((train_pred - train_data$AccidentCount)^2))
train_r2 <- cor(train_pred, train_data$AccidentCount)^2

cat("Training RMSE:", train_rmse, "\n")
cat("Training R-squared:", train_r2, "\n")

# 7. Test evaluation (optional)
test_pred <- predict(rf_model, test_data)
test_rmse <- sqrt(mean((test_pred - test_data$AccidentCount)^2))
test_r2 <- cor(test_pred, test_data$AccidentCount)^2

cat("Test RMSE:", test_rmse, "\n")
cat("Test R-squared:", test_r2, "\n")

# 8. Model output
print(rf_model)
varImpPlot(rf_model)
```

Attempt 2:\

```{r}
run_casualty_ridge <- function(data, alpha = 0, lambda = NULL, exclude_cols = NULL, seed = NULL){
  #' Run Ridge Regression for Number_of_Casualties prediction
  #'
  #' @param data Input data frame
  #' @param alpha 0 for ridge (L2), 1 for lasso (L1)
  #' @param lambda Regularization strength (if NULL, cv.glmnet will choose)
  #' @param exclude_cols Vector of column names to exclude from predictors
  #' @param seed Optional random seed for reproducibility
  library(glmnet)
  library(dplyr)
  
  # Set seed if provided
  if (!is.null(seed)) set.seed(seed)
  
  # Fixed parameters
  sample_fraction <- 0.01
  train_frac <- 0.7
  target_var <- "Number_of_Casualties"
  
  # Sample data
  reduced_data <- data[sample(nrow(data), size = round(sample_fraction * nrow(data))), ]
  
  # Split into train/test
  index <- sample(1:nrow(reduced_data), round(nrow(reduced_data) * train_frac))
  train <- reduced_data[index, ]
  test <- reduced_data[-index, ]
  
  # Remove excluded columns and prepare matrices
  if (!is.null(exclude_cols)) {
    predictors <- setdiff(names(train), c(target_var, exclude_cols))
  } else {
    predictors <- setdiff(names(train), target_var)
  }
  
  x_train <- model.matrix(~ . - 1, train[, predictors])  # -1 removes intercept
  y_train <- train[[target_var]]
  x_test <- model.matrix(~ . - 1, test[, predictors])
  y_test <- test[[target_var]]
  
  # Cross-validated Ridge Regression
  if (is.null(lambda)) {
    cv_fit <- cv.glmnet(x_train, y_train, alpha = alpha)
    lambda <- cv_fit$lambda.min
  }
  
  ridge_model <- glmnet(x_train, y_train, alpha = alpha, lambda = lambda)
  
  # Generate predictions
  test_predictions <- predict(ridge_model, newx = x_test)
  
  # Calculate metrics
  rmse <- sqrt(mean((test_predictions - y_test)^2))
  r_squared <- cor(test_predictions, y_test)^2
  
  # Get variable importance (absolute coefficient values)
  coefs <- coef(ridge_model)[-1, ]  # Remove intercept
  importance <- data.frame(
    variable = names(coefs),
    importance = abs(coefs),
    row.names = NULL
  ) %>% arrange(desc(importance))
  
  # Plot coefficients
  plot(ridge_model, xvar = "lambda", label = TRUE)
  title("Ridge Regression Coefficients", line = 2.5)
  
  # Return results
  return(list(
    model = ridge_model,
    cv_fit = if(exists("cv_fit")) cv_fit else NULL,
    train = train,
    test = test,
    predictions = test_predictions,
    rmse = rmse,
    r_squared = r_squared,
    importance = importance,
    lambda = lambda,
    formula = paste(target_var, "~", paste(predictors, collapse = " + "))
  ))
}
```

```{r}

results <- run_casualty_ridge(data,exclude_cols = c("Accident_Severity", "Road_Type", "Speed_limit", "Junction_Control","Light_Conditions", "Weather_Conditions", "Road_Surface_Conditions"))



# Get top 5 important variables
head(results$importance, 5)


```
